---
title: "Version-controlled data with doltliter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Version-controlled data with doltliter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette walks through a full cycle: connect, write a table, commit,
branch, modify, diff, merge, and resolve a conflict.

```{r setup}
library(doltliter)
```

## Connecting

A DoltLite database is a single file.

```{r}
path <- tempfile(fileext = ".db")
con <- DBI::dbConnect(Doltlite(), path)
con
```

One caveat worth stating up front: pass a real path, not `""`. SQLite's
anonymous temporary database is created in the *original* B-tree format, and
version control is not available there.

Commits are attributed per connection, and the setting is not persisted, so
configure it after connecting:

```{r}
dolt_config(con, user.name = "Ada Lovelace", user.email = "ada@example.com")
```

## Writing data

Nothing surprising: this is ordinary DBI.

```{r}
users <- data.frame(
  id     = 1:3,
  name   = c("ada", "bob", "cleo"),
  active = c(1L, 0L, 1L),
  stringsAsFactors = FALSE
)

DBI::dbWriteTable(con, "users", users,
                  field.types = c(id = "INTEGER PRIMARY KEY"))
DBI::dbReadTable(con, "users")
```

The primary key is declared here because `dolt_blame()` later needs one: row
attribution has to have a stable row identity to attribute to. An
`INTEGER PRIMARY KEY` stays an ordinary writable `rowid` alias, exactly as in
SQLite. A *non*-`INTEGER` primary key is different — it makes the table
clustered, with no `rowid` at all and `NOT NULL` key columns.

The data is in the database, but nothing is in the history yet:

```{r}
dolt_status(con)
```

## The two kinds of commit

This is the distinction to hold on to.

`DBI::dbBegin()` / `DBI::dbCommit()` delimit a **SQL transaction**: the
statements between them either all apply or none do. That is about atomicity.

`dolt_commit()` writes a **commit into the version history**, the way
`git commit` does. That is about being able to come back later.

`dbCommit()` is "save the file"; `dolt_commit()` is "commit to the
repository". Writing a table and committing the SQL transaction leaves the
rows in `dolt_status()` until you make a Dolt commit:

```{r}
dolt_commit(con, "Initial load")
dolt_status(con)
dolt_log(con)[, c("committer", "message")]
```

A fresh database already contains one commit — DoltLite's own
`Initialize data repository`.

## Branching

Each connection has its own active branch.

```{r}
dolt_branch(con, "experiment")
dolt_checkout(con, "experiment")
active_branch(con)
```

Make a change on the branch and commit it:

```{r}
DBI::dbExecute(con, "UPDATE users SET active = 1 WHERE id = 2")
dolt_commit(con, "Activate bob")
```

`main` is untouched:

```{r}
dolt_checkout(con, "main")
DBI::dbGetQuery(con, "SELECT id, name, active FROM users ORDER BY id")
```

You can also connect straight onto a branch. `doltliter` translates this to
DoltLite's `dbname@branch` path syntax internally, and errors if the branch
does not exist rather than silently opening `main`:

```{r}
con2 <- DBI::dbConnect(Doltlite(), path, branch = "experiment")
active_branch(con2)
DBI::dbDisconnect(con2)
```

## Diffing

`dolt_table_diff()` compares a table between two refs, with `from_`/`to_`
column pairs:

```{r}
d <- dolt_table_diff(con, "users", from = "main", to = "experiment")
d[, c("diff_type", "to_id", "from_active", "to_active")]
```

`dolt_diff()` works at the level of *which tables* changed in each commit. The
pseudo-commit `WORKING` covers uncommitted changes:

```{r}
DBI::dbExecute(con, "UPDATE users SET name = 'ADA' WHERE id = 1")
dolt_diff(con, "users")[, c("commit_hash", "table_name", "data_change")]
dolt_reset(con, "hard")
```

For summaries rather than rows, there are `dolt_diff_stat()`,
`dolt_diff_summary()` and `dolt_schema_diff()`; for an executable rebuild,
`dolt_patch()`.

## Merging

A clean merge just works:

```{r}
dolt_merge(con, "experiment")
DBI::dbGetQuery(con, "SELECT id, name, active FROM users ORDER BY id")
```

## Resolving a conflict

Conflicts in DoltLite are **never durable**: they exist only inside the
transaction that produced them, and a commit is refused while any remain. An
autocommit merge that conflicts is therefore rolled back whole — so to inspect
and resolve conflicts, run the merge inside a transaction.

Set up two branches that edit the same row:

```{r}
dolt_checkout(con, "theirs", create = TRUE)
DBI::dbExecute(con, "UPDATE users SET name = 'THEIRS' WHERE id = 1")
dolt_commit(con, "Rename on theirs")

dolt_checkout(con, "main")
DBI::dbExecute(con, "UPDATE users SET name = 'OURS' WHERE id = 1")
dolt_commit(con, "Rename on main")
```

Now merge inside a transaction:

```{r}
DBI::dbBegin(con)
dolt_merge(con, "theirs")
dolt_conflicts(con)
```

Inspect the detail, then take one side (or edit the rows directly):

```{r}
dolt_conflicts_resolve(con, "ours")
dolt_conflicts(con)
dolt_commit(con, "Merge theirs, keeping ours")
DBI::dbGetQuery(con, "SELECT id, name FROM users WHERE id = 1")
```

`dolt_commit()` also ends the SQL transaction, so there is nothing left to
`dbCommit()`.

## History and blame

```{r}
dolt_log(con)[, c("committer", "message")]
```

`dolt_history()` gives every version of every row, and `dolt_at()` reads a
table as it was at a revision:

```{r}
first <- rev(dolt_log(con)$commit_hash)[1]
dolt_at(con, "users", first)
```

`dolt_blame()` attributes each live row to the commit that last set it. It
needs a primary key — there is no stable row identity without one:

```{r}
dolt_blame(con, "users")[, c("id", "committer", "message")]
```

## dplyr

`dplyr::tbl()` works over the connection with no extra setup: DoltLite's SQL
dialect is SQLite's, and `doltliter` reuses dbplyr's SQLite translation.

```{r, eval = requireNamespace("dplyr", quietly = TRUE)}
library(dplyr)
tbl(con, "users") |>
  filter(active == 1) |>
  select(id, name) |>
  collect()
```

## Remotes

Databases can push, pull, fetch and clone between each other, over the
filesystem or HTTP:

```{r}
origin <- tempfile(fileext = ".db")
file.copy(path, origin)

clone_path <- tempfile(fileext = ".db")
clone <- DBI::dbConnect(Doltlite(), clone_path)
dolt_clone(clone, paste0("file://", origin))

DBI::dbReadTable(clone, "users")
dolt_remotes(clone)[, c("name", "url")]
DBI::dbDisconnect(clone)
```

## Concurrency

DoltLite allows one durable writer at a time. A second connection that begins
a write transaction gets `SQLITE_BUSY` until the first finishes, and a
transaction that tries to upgrade after a peer advanced the store gets
`SQLITE_BUSY_SNAPSHOT`. Both are retryable, and `dbConnect()` sets a
five-second busy timeout by default:

```{r, eval = FALSE}
con <- DBI::dbConnect(Doltlite(), path, busy_timeout = 30000)
```

```{r}
DBI::dbDisconnect(con)
```
