---
title: "Using dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_dplyr <- requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("dbplyr", quietly = TRUE)
knitr::opts_chunk$set(eval = has_dplyr)
```

`dplyr::tbl()` works over a `doltliter` connection with no extra setup.
DoltLite's SQL dialect *is* SQLite's — the parser and planner are upstream, only
the storage engine below them differs — so this package reuses dbplyr's SQLite
translation rather than defining its own.

```{r setup, message = FALSE}
library(doltliter)
library(dplyr)

con <- DBI::dbConnect(Doltlite(), tempfile(fileext = ".db"))
dolt_config(con, user.name = "Ada Lovelace", user.email = "ada@example.com")
```

## A lazy table

```{r}
measurements <- data.frame(
  id     = 1:8,
  site   = rep(c("north", "south"), each = 4),
  season = rep(c("spring", "summer"), times = 4),
  value  = c(12.1, 15.4, 9.8, 11.2, 18.7, 22.3, 17.1, 19.9)
)

copy_to(con, measurements, "measurements", temporary = FALSE)
dolt_commit(con, "Load measurements")

tb <- tbl(con, "measurements")
tb
```

The pipeline is not evaluated until you ask for the rows:

```{r}
tb |>
  filter(value > 15) |>
  select(site, season, value) |>
  arrange(desc(value)) |>
  collect()
```

`show_query()` prints the SQL that would run, which is the quickest way to
check that a step is being pushed down rather than pulled into R:

```{r}
tb |>
  group_by(site) |>
  summarise(n = n(), mean_value = mean(value, na.rm = TRUE)) |>
  show_query()
```

```{r}
tb |>
  group_by(site) |>
  summarise(n = n(), mean_value = mean(value, na.rm = TRUE)) |>
  collect()
```

## Comparing branches with dplyr

This is where the combination gets interesting. A branch is just a different
view of the same file, so you can point two connections at two branches and
compare them with ordinary dplyr.

```{r}
path <- DBI::dbGetInfo(con)$dbname

dolt_checkout(con, "reprocessed", create = TRUE)
DBI::dbExecute(con, "UPDATE measurements SET value = value * 1.1 WHERE site = 'north'")
dolt_commit(con, "Apply 10% correction to northern sites")
DBI::dbDisconnect(con)
```

Open one connection per branch:

```{r}
main <- DBI::dbConnect(Doltlite(), path, branch = "main")
repro <- DBI::dbConnect(Doltlite(), path, branch = "reprocessed")

before <- tbl(main,  "measurements") |>
  group_by(site) |> summarise(mean_value = mean(value, na.rm = TRUE)) |> collect()

after <- tbl(repro, "measurements") |>
  group_by(site) |> summarise(mean_value = mean(value, na.rm = TRUE)) |> collect()

inner_join(before, after, by = "site", suffix = c("_main", "_reprocessed"))
```

Note that a join *across* two connections happens in R, because the two
`tbl()`s belong to different sources. Within one connection, joins are pushed
down to SQL as usual.

## Writing results back

`compute()` materialises a query into a table on the database, which then
commits, diffs and branches like any other:

```{r}
summary_tbl <- tbl(repro, "measurements") |>
  group_by(site, season) |>
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") |>
  compute(name = "site_season_summary", temporary = FALSE)

dolt_config(repro, user.name = "Ada Lovelace", user.email = "ada@example.com")
dolt_commit(repro, "Add site/season summary")
dolt_log(repro)$message[1]
```

## A note on translation

dbplyr version-gates a handful of its SQLite translations and reads the SQLite
version from RSQLite. If RSQLite is not installed, `doltliter` falls back to
dbplyr's default translation and says so once. Everything still works; you just
get slightly more generic SQL. Install RSQLite if you want the full SQLite
translation table.

```{r}
DBI::dbDisconnect(main)
DBI::dbDisconnect(repro)
```
