---
title: "Installation and linking strategies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation and linking strategies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`doltliter` embeds a C library. Unlike RSQLite, which vendors `sqlite3.c` and
compiles it unconditionally, this package has to *find* a `libdoltlite` to
build against — and the right way to do that differs by platform. This article
explains what `configure` does, so that when it picks something unexpected you
know why.

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("CathalByrneGit/doltliter")
```

## The three strategies

`configure` tries these in order of cost, and prints which one it chose.

### 1. An installed DoltLite (`system`)

Costs nothing if you already have DoltLite. `configure` looks at
`pkg-config`, then `DOLTLITE_HOME`, then explicit `DOLTLITE_CFLAGS` /
`DOLTLITE_LIBS`, then the usual prefixes (`/usr/local`, `/usr`,
`/opt/homebrew`, `/opt/local`).

```sh
sudo bash -c 'curl -fsSL https://github.com/dolthub/doltlite/releases/latest/download/install.sh | bash'
# Debian/Ubuntu alternative: install libdoltlite-dev
```

### 2. A vendored amalgamation (`vendor`)

DoltLite publishes a single-file amalgamation, the same shape as SQLite's.
Staging it means the package compiles with **no network access at install
time**, which is the form a CRAN submission would take, and pins the DoltLite
version exactly.

```sh
Rscript tools/vendor_amalgamation.R
R CMD INSTALL .
```

It costs about a minute of compile time. This is the default on Windows and
Intel macOS, for reasons in the next section.

### 3. A downloaded release library (`download`)

`configure` fetches the prebuilt library matching your OS and architecture
from DoltLite's GitHub releases. Best for a one-line `install_github()` on a
machine with no system install.

### Forcing one

```sh
DOLTLITER_STRATEGY=system   R CMD INSTALL .   # or vendor, download
DOLTLITE_VERSION=0.50.3     R CMD INSTALL .   # pin the DoltLite version
```

## Why Windows and Intel macOS vendor

This is not arbitrary. Checking the actual release artifacts turned up two
things the upstream README does not mention:

* A Windows library archive **does** exist — but it contains only
  `libdoltlite.dll`, with no static archive and no `.lib` / `.dll.a` import
  library. There is nothing for Rtools to link against, so a prebuilt library
  is not usable there.
* **No Intel macOS library is published at all.** `install.sh` lists `osx-x64`
  among its targets, but no such asset exists.

On both platforms the amalgamation is the only route that works, so
`configure` reaches for it first.

## Checking what you got

`configure` finishes by compiling and running a probe that asserts the engine
really is DoltLite's prolly tree, so a build that accidentally linked stock
SQLite fails at install time rather than at your first `dolt_*` call. You can
re-check at any point:

```{r}
library(doltliter)

doltlite_engine()    # "prolly" for a DoltLite build
doltlite_version()   # the bundled SQLite version
```

`dolt_version()` needs a connection, and reports the DoltLite release:

```{r}
con <- DBI::dbConnect(Doltlite(), tempfile(fileext = ".db"))
dolt_version(con)
DBI::dbGetInfo(con)[c("doltlite.engine", "doltlite.version", "branch")]
DBI::dbDisconnect(con)
```

## Two things that will surprise you

**An empty `dbname` is not version controlled.** SQLite's anonymous temporary
database is created in the *original* B-tree format, so none of the `dolt_*`
functions exist there. The package says so plainly rather than letting SQLite
report "no such function":

```{r, error = TRUE}
tmp <- DBI::dbConnect(Doltlite(), "")
DBI::dbGetQuery(tmp, "SELECT doltlite_engine() AS engine")
dolt_status(tmp)
```

```{r, include = FALSE}
DBI::dbDisconnect(tmp)
```

Use a real path — `tempfile()` is fine — whenever you want version control.

**The installed package is large.** Around 19 MB, because the engine is
statically linked into the shared library. That is inherent to embedding
DoltLite, and the same trade RSQLite makes with `sqlite3.c`.

## If the build fails

`configure` writes everything it tried to `config.log` in the package root.
That file names the compiler invocation, the candidate link lines and which
one succeeded, and is the first place to look. The error message also lists
the three remedies above.
