---
title: "Time travel: history, diffs and blame"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time travel: history, diffs and blame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Committing data is only half the point. The other half is being able to ask
what the data used to be, what changed between two moments, and who changed it.

```{r setup}
library(doltliter)

con <- DBI::dbConnect(Doltlite(), tempfile(fileext = ".db"))
dolt_config(con, user.name = "Ada Lovelace", user.email = "ada@example.com")
```

Let's build a table with a few revisions behind it. The primary key matters
here: `dolt_blame()` needs one, because without a stable row identity there is
nothing to attribute a change *to*.

```{r}
DBI::dbWriteTable(
  con, "readings",
  data.frame(id = 1:3,
             sensor = c("A", "B", "C"),
             value  = c(10.0, 20.0, 30.0)),
  field.types = c(id = "INTEGER PRIMARY KEY")
)
dolt_commit(con, "First calibration run")

DBI::dbExecute(con, "UPDATE readings SET value = 21.5 WHERE id = 2")
dolt_commit(con, "Recalibrate sensor B")

DBI::dbExecute(con, "INSERT INTO readings VALUES (4, 'D', 40.0)")
dolt_commit(con, "Add sensor D")
```

## What happened, and when

```{r}
dolt_log(con)[, c("committer", "date", "message")]
```

`dolt_diff()` works one level up — which *tables* changed in each commit,
rather than which rows:

```{r}
dolt_diff(con, "readings")[, c("commit_hash", "data_change", "schema_change")]
```

Uncommitted work shows up under the pseudo-commit `WORKING`, which is a
convenient way to see whether anything is pending:

```{r}
DBI::dbExecute(con, "UPDATE readings SET value = 41.0 WHERE id = 4")
dolt_diff(con, "readings")$commit_hash[1]
dolt_status(con)
```

## Row-level diffs

`dolt_table_diff()` is the workhorse. Give it two endpoints and it returns one
row per changed row, with `from_` and `to_` columns side by side:

```{r}
d <- dolt_table_diff(con, "readings", from = "HEAD", to = "WORKING")
d[, c("diff_type", "to_id", "from_value", "to_value")]
```

Endpoints are anything DoltLite understands as a ref — a branch, a tag, a
commit hash, `HEAD~2`, `WORKING`:

```{r}
dolt_table_diff(con, "readings", from = "HEAD~2", to = "HEAD")[
  , c("diff_type", "to_id", "from_value", "to_value")]
```

A range string works too, which is handy when comparing branches.
Two dots are the two endpoints; three dots is merge-base to the right endpoint:

```{r, eval = FALSE}
dolt_table_diff(con, "readings", range = "main..feature")
dolt_table_diff(con, "readings", range = "main...feature")
```

For a summary rather than the rows themselves:

```{r}
dolt_diff_stat(con, "HEAD~2", "HEAD")
```

```{r}
dolt_reset(con, "hard")
```

## Reading the past directly

`dolt_at()` reads a table as it was at some revision — no checkout, no
branch switching, just a query against a historical snapshot:

```{r}
first <- rev(dolt_log(con)$commit_hash)[2]   # the first real commit

dolt_at(con, "readings", first)
DBI::dbReadTable(con, "readings")
```

`dolt_history()` goes wider: every version of every row across the commit
ancestry, with the commit metadata attached. It is the right tool for "plot
this value over time":

```{r}
h <- dolt_history(con, "readings")
h[h$id == 2, c("id", "sensor", "value", "commit_hash")]
```

## Who changed this row

`dolt_blame()` attributes each *live* row to the commit that last set it:

```{r}
dolt_blame(con, "readings")[, c("id", "committer", "message")]
```

It walks first parents from `HEAD`. Schema-only changes such as
`ALTER TABLE ... ADD COLUMN` do not update blame, so a column addition will not
make every row look freshly touched.

## Content hashes

Every table root and the catalog as a whole are content-addressed, and the
table and database hashes are *history-independent*: the same rows hash the
same regardless of how they got there.

```{r}
dolt_hashof(con, "HEAD")        # commit hash — covers the timestamp too
dolt_hashof_table(con, "readings")
dolt_hashof_db(con)
```

That makes `dolt_hashof_table()` a cheap way to ask "did this table actually
change?" across two databases or two points in time, without diffing.

## Generating a patch

`dolt_patch()` turns a diff into ordered, executable SQLite statements — useful
for review, or for replaying a change somewhere else:

```{r}
p <- dolt_patch(con, "HEAD~1", "HEAD")
p[, c("statement_order", "diff_type", "statement")]
```

```{r}
DBI::dbDisconnect(con)
```
