---
title: "Branching, merging and conflicts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Branching, merging and conflicts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`vignette("doltliter")` covers a single pass through the commit loop. This
article is about working on more than one line of history at once: branching,
merging, what happens when two branches disagree, and how to undo things.

```{r setup}
library(doltliter)

path <- tempfile(fileext = ".db")
con <- DBI::dbConnect(Doltlite(), path)
dolt_config(con, user.name = "Ada Lovelace", user.email = "ada@example.com")
```

We will curate a small reference table — the kind of data that gets corrected
over time, which is where version control earns its keep.

```{r}
sites <- data.frame(
  id     = 1:4,
  name   = c("Ballycotton", "Roches Point", "Mizen Head", "Loop Head"),
  region = c("Cork", "Cork", "Cork", "Clare"),
  active = c(1L, 1L, 0L, 1L),
  stringsAsFactors = FALSE
)

DBI::dbWriteTable(con, "sites", sites,
                  field.types = c(id = "INTEGER PRIMARY KEY"))
dolt_commit(con, "Initial site list")
```

## Branches are per connection

Each connection holds its own active branch. Creating one does not move you
onto it:

```{r}
dolt_branch(con, "review")
dolt_branches(con)[, c("name", "dirty")]
active_branch(con)
```

`dolt_checkout()` switches; `create = TRUE` does both at once.

```{r}
dolt_checkout(con, "review")
active_branch(con)
```

Work done here does not touch `main`:

```{r}
DBI::dbExecute(con, "UPDATE sites SET active = 1 WHERE name = 'Mizen Head'")
dolt_commit(con, "Reactivate Mizen Head")

dolt_checkout(con, "main")
DBI::dbGetQuery(con, "SELECT name, active FROM sites WHERE name = 'Mizen Head'")
```

One thing to be aware of: uncommitted work belongs to the *branch*, not the
connection. There is no stash — checking out does not shelve changes, and
another connection that checks out the same branch will see the same working
set.

## A clean merge

When the two sides touch different rows, the merge just happens and returns
the new commit hash:

```{r}
dolt_merge(con, "review")
DBI::dbGetQuery(con, "SELECT name, active FROM sites ORDER BY id")
```

## A conflicting merge

Now the interesting case. Two branches edit the *same* row:

```{r}
dolt_checkout(con, "regional", create = TRUE)
DBI::dbExecute(con, "UPDATE sites SET region = 'County Cork' WHERE id = 1")
dolt_commit(con, "Use full county names")

dolt_checkout(con, "main")
DBI::dbExecute(con, "UPDATE sites SET region = 'CORK' WHERE id = 1")
dolt_commit(con, "Normalise regions to upper case")
```

Here is the part that catches people out. Conflicts in DoltLite are **never
durable** — they exist only inside the transaction that produced them. A merge
run in autocommit mode, which is what you get by default, is rolled back in
full the moment it conflicts:

```{r, error = TRUE}
dolt_merge(con, "regional")
```

Nothing was left behind to inspect:

```{r}
nrow(dolt_conflicts(con))
DBI::dbGetQuery(con, "SELECT region FROM sites WHERE id = 1")
```

To actually work on the conflict, put the merge in a transaction:

```{r}
DBI::dbBegin(con)
dolt_merge(con, "regional")
dolt_conflicts(con)
```

The per-table view carries `base_`, `our_` and `their_` columns, so you can see
all three sides at once:

```{r}
detail <- dolt_conflicts_table(con, "sites")
detail[, grep("region|dolt_conflict_id", names(detail), value = TRUE)]
```

Resolve by taking one side wholesale, or by editing the rows directly and then
resolving:

```{r}
dolt_conflicts_resolve(con, "ours")
nrow(dolt_conflicts(con))

dolt_commit(con, "Merge regional, keeping upper-case regions")
DBI::dbGetQuery(con, "SELECT region FROM sites WHERE id = 1")
```

Note that `dolt_commit()` also ended the SQL transaction, so there is nothing
left for `DBI::dbCommit()` to do:

```{r, error = TRUE}
DBI::dbCommit(con)
```

A commit is refused while any conflict remains, so you cannot accidentally
record a half-merged state.

## Tags

Tags name a commit so you can come back to it:

```{r}
dolt_tag(con, "v1.0")
dolt_tags(con)[, 1:2]
```

They work anywhere a ref does — `dolt_at(con, "sites", "v1.0")`, or as a diff
endpoint.

## Undoing things

`dolt_reset()` throws away uncommitted work; `"soft"` unstages but keeps it,
`"hard"` discards it:

```{r}
DBI::dbExecute(con, "DELETE FROM sites WHERE id = 4")
DBI::dbGetQuery(con, "SELECT count(*) AS n FROM sites")$n

dolt_reset(con, "hard")
DBI::dbGetQuery(con, "SELECT count(*) AS n FROM sites")$n
```

`dolt_revert()` is the one to reach for when the change is already committed.
It does not rewrite history — it writes a *new* commit applying the inverse:

```{r}
DBI::dbExecute(con, "INSERT INTO sites VALUES (5, 'Blackrock', 'CORK', 1)")
oops <- dolt_commit(con, "Add Blackrock")

dolt_revert(con, oops)
DBI::dbGetQuery(con, "SELECT count(*) AS n FROM sites")$n
dolt_log(con)$message[1:2]
```

`dolt_cherry_pick()` applies a single commit from elsewhere onto the current
branch, and `dolt_rebase()` replays a whole branch onto another. Both resolve
conflicts the same way `dolt_merge()` does, so the transaction advice above
applies to them too.

```{r}
DBI::dbDisconnect(con)
```
