---
title: "The pane"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`dolt_pane()` is a Shiny gadget, and an RStudio addin, in the spirit of
RStudio's Git pane: what has changed, what the change looks like row by row,
the history, and the branch operations — over a connection you already have
open.

```{r}
library(doltliter)

con <- DBI::dbConnect(Doltlite(), "mydata.db")
dolt_pane(con)
```

Call it with no arguments and it uses the single open connection it can find;
if there are several, it says so rather than guessing. There is also an
**Addins → DoltLite pane** entry, which does the same thing.

It needs the suggested packages:

```{r}
install.packages(c("shiny", "miniUI", "DT"))
```

## It runs on your connection, and that is the whole design

The obvious way to build this would be a background process watching the
database file, so the pane could sit there permanently like the Git pane does.
That does not work here, and it is worth knowing why before the screenshots
make it look simpler than it is.

A background pane means a *second connection*. Measured against a first:

| What a pane needs | On a second connection |
|---|---|
| See uncommitted changes | works |
| Commit them | works |
| Report the branch you are on | **wrong** — you on `feature`, it says `main` |
| See work inside an open transaction | **blind** — nothing, while you have changes |
| Write while you hold a transaction | **fails** — `database is locked` |

The third row is the one that settles it. Branches in DoltLite are per
connection, so a background pane would show its own branch rather than yours,
and pressing Commit could land your work on a branch you are not on, silently.
Rows four and five are each independently fatal.

So the gadget runs in your R session, on the connection you hand it. The cost
is that it blocks the console while open, which is the honest trade rather than
a limitation to apologise for: the state that matters — your branch, your
transaction, your uncommitted rows — lives in your connection and nowhere else.

## Changes

The working set, the row-level diff for whichever table you select, and a
commit box.

![The Changes tab: two modified tables, the diff for one of them, and the
commit box](pane-changes.png)

The diff is where this beats the Git pane rather than imitating it. Git's
diffs are text, so a GUI adds modest value. `dolt_table_diff()` returns
*structured rows* — `from_`/`to_` column pairs — which is genuinely better in a
sortable table than printed at the console, and it is the thing you most want
to look at before committing. `diff_type` leads because it is what you scan
for, and is coloured green for added, amber for modified, red for removed.

Two things the buttons are explicit about:

* **Commit refuses while a SQL transaction is open.** `dolt_commit()` ends the
  enclosing transaction, so committing from the gadget mid-`dbBegin()` would
  silently end a transaction your console still believes is running.
* **"Stage everything and commit"** is named that way because `dolt_add("-A")`
  sweeps the whole working set, not just the table on screen.

## Branches

List, create, switch, and merge.

![The Branches tab: the branch list with the current branch marked, and
controls to create, switch and merge](pane-branches.png)

## Merging, and conflicts

Merges run inside a transaction because they have to. DoltLite rolls a
conflicting merge back whole in autocommit mode, so there would be nothing left
to inspect — conflicts are only reachable inside `BEGIN`/`COMMIT`. When one
conflicts, the gadget holds that transaction open and shows you base, ours and
theirs side by side.

![A conflicted merge: the conflicted tables listed with their counts, and base,
ours and theirs for the selected one](pane-conflicts.png)

Conflicts are handled one table at a time, because the conflict columns follow
the shape of the table they belong to — `our_team` for one table,
`our_region` for another — so several conflicted tables genuinely cannot share
a view. The picker lists them with their counts and the detail follows your
selection.

`Keep ours` and `Keep theirs` resolve the table you are looking at, not every
conflicted one. That is deliberate twice over: a button that silently acted on
tables off screen would be a trap, and the whole reason two tables conflict
separately is that they may deserve different answers.

![After resolving one table: it leaves the list, the other stays
conflicted](pane-resolving.png)

Tables leave the list as they are resolved, so it empties as you work through
it. Then `Commit merge` finishes, or `Abort` rolls the whole thing back —
which restores the rows and clears the merge, since a rollback is exactly what
undoes an unfinished merge.

Two consequences worth keeping in mind:

* **A conflicted merge means an open transaction on your connection.** Finish
  or abort it before going back to the console. Closing the window rolls it
  back rather than stranding it.
* **A merge is refused while the working set is dirty**, because DoltLite
  requires a clean one. Commit on the Changes tab first.

## What is not there

**Remotes.** `dolt_push()`, `dolt_pull()` and friends are not wrapped here.
They need a configured remote and credentials, and their side effects reach
beyond your machine — a different class of risk from everything above.

**Tags, reset, revert, cherry-pick.** All available as functions; none have
buttons yet.

## Packaging

`shiny`, `miniUI`, `DT` and `rstudioapi` are `Suggests`, not `Imports`. This is
one optional entry point in a package whose job is to be a DBI backend, and
most users of the backend never open it. `dolt_pane()` names what is missing
rather than failing obscurely:

```{r}
dolt_pane(con)
#> Error: dolt_pane() needs shiny, miniUI and DT. Install with:
#>   install.packages(c("shiny", "miniUI", "DT"))
```

Four suggested packages for one function is a fair argument for eventually
moving the pane into a companion package. It costs nothing while the package
is distributed from GitHub.
